---
title: "latimesproject"
author: "Heimbrock"
date: "2025-10-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
library(tidyverse)
library(janitor)
library(lubridate)
```

#I am documenting my code and loading libaries#
```{r}
library(tidyverse)
```

#Loading data#

```{r}
payouts<-rio::import('/Users/rebeccaheimbrock/Desktop/jour772/la-settlements-analysis/input/payouts.xlsx', skip=1, header=TRUE)
#David note: skip/delete the wonky top row, make the second row into headers #Bex note: I got this code from David after fighting with Claude.
  payouts<-payouts %>% 
  slice_head(n=18701) %>% 
  clean_names()
```
#Goal: get code to do this, somehow .. "Last budget year, the city paid out more than $200 million in legal settlements and court judgments — a record amount that was more than the city spent on its libraries or fixing its streets."

```{r}
yearly_data_sum <- payouts |> 
  group_by(fiscal_year) |>
  summarise(
    total_cases = n(),
    total_amount = sum(amount, na.rm = TRUE),
  ) |> 
  arrange(fiscal_year)
#Here was my original code, which did everything I wanted except combine all instances of X year to make one data column per year: 
#year_amount <- payouts |> 
#group_by(fiscal_year, amount) |> 
#summarise(total_amount=sum(amount)) |> 
#arrange(desc(amount))
#That didn't work, so...
#I looked back through my R notebook and found only stringer detect functions.
#This is what I asked Claude: How to combine all instances of a year being mentioned in a spreadsheet? Claude then asked me to explain whether the data was in its own column, and how it was formatted.
#Claude suggested the above code with the following explanatory comment: #total_cases: number of payments made that year
#total_amount: sum of all payments
#avg_amount: average payment amount
#min_amount and max_amount: smallest and largest payments
#Thanks, Claude, but that's kind of a lot - I just want to know the total amount. So, all I included was total_amount.
```

#So, the total for 2018 is $72,829,597... that doesn't seem right. LA Times says the city paid out $200million! But, the total for 2017 is over $201million. This article was published in 2018. So, last budget year was, in fact, 2017. Yay!
#Now it's time to read in the csv file containing totals spent on public works.
```{r}
dept_totals<-rio::import('/Users/rebeccaheimbrock/Desktop/jour772/la-settlements-analysis/output/department-totals.csv')
```
#Hey R, how much was spent on public works and libraries? 
```{r}
cherry_picking_data_for_maximum_effect <- dept_totals |>
  filter(str_detect(latimes_department, 'PUBLIC WORKS|LIBRARY')) |>
  arrange(desc(amount))
```
#This was my initial code...
#> cherry_picking_data_for_maximum_effect <- dept_totals |>
#filter(str_detect(latimes_department, 'PUBLIC WORKS', 'LIBRARY))

#That didn't work. I went back to pre-lab03 and tried to string together that code, this is what I ended up with...
#> cherry_picking_data_for_maximum_effect <- dept_totals |>
#sort_by(amount) |>
#filter(str_detect('PUBLIC WORKS', 'LIBRARY')) |>
#arrange(desc(amount))

#That also didn't work. So then I asked Claude... 
#Why isn't this code working: cherry_picking_data_for_maximum_effect <- dept_totals |>
# filter(str_detect('PUBLIC WORKS', 'LIBRARY')) |>
# arrange(desc(amount))

#Claude told me to fix three things...
#First argument should be the column name (like department) - not in quotes
#Use the | (OR) operator to find either "PUBLIC WORKS" or "LIBRARY"
#Make sure you're using the correct column name from your dataframe

#So that's how I got the above data.
