David Landerman---
title: "la_reverse_engineering"
output: html_document
date: "2025-10-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#option_command_i

#Loading Libaries
```{r}
library(tidyverse)
library(rio)
library(lubridate)
```

#control-click-hold option, copy path

#'/Users/davidlanderman/UMD Drive/Github/jour772/project/reverse engineering/la-settlements-analysis/input/payouts.xlsx'

```{r}
payouts<-rio::import('/Users/davidlanderman/UMD Drive/Github/reverse engineering/la-settlements-analysis/input/payouts.xlsx', skip=1, header=TRUE)
#skip/delete the wonky top row, make the second row into headers
  payouts<-payouts %>% 
    #the bottom two rows are a space and then a grand total, this will screw up the data. Slice_head means filtering by that number of rows, I tried to use slice_tail(n=2) but it gave me JUST the bottom two rows, so I did total rows minus 2 and slice_head.
    slice_head(n=18701)
```
AI USE DISCLOSURE
Googled "is there a way to import a csv into r and delete the top row and use row 2 as the header"
The first link was substack, read it, not helpful but the top result was google's ai which said this
"Yes, it is possible to import a CSV file into R, skip the first row, and use the second row as the header. This can be achieved using the read.csv() function from base R or the read_csv() function from the readr package.
Using read.csv() (Base R):
Code

data <- read.csv("your_file.csv", skip = 1, header = TRUE)"




